---
title: "SDMs: Honours Guide"
author: "Clarke van Steenderen"
date: "2025"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

## First steps

This introductory module to Species Distribution Modelling (SDM) will guide you through all the steps needed to create your first prediction models using the popular MaxEnt tool through R. We will use the Asian Citrus Psyllid (*Diaphorina citri*) as a case study, and generate a predictive MaxEnt model based on global climate variables. *Diaphorina citri* is a major citrus pest and has caused significant financial losses to citrus industries in many regions; in some cases it has threatened their existence. The psyllid is native to Asia, but has spread across the world. It has been recorded in Africa, but it has not dispersed as far south as South Africa yet.

Let's create a model to predict which regions of South Africa might be the most climatically suitable for *D. citri*, and how it might enter the country through climatically-suitable routes.

🟣 Create a new R project in a folder on your PC, and run the code below bit by bit. You can save this all in one big script, or you can create multiple R scripts for each part of the process. You can use the command ```source("script_name.R")``` to read in a particular script into your R environment.

🟣 All data required is available here: **https://drive.google.com/drive/folders/1qe3cojnh71l2tmwR4WsdlGPr4fR7CHf0?usp=drive_link**
Save this data into the same directory as your R project.

🟢 There are parts of the code that are for you to complete - these are marked by a green dot: 🟢

### Install/load the required R packages, and set the aesthetic theme for the plots

These are the R libraries required to run the various functions we will need. If these are not installed, use ```install.packages("library_name")``` to save and access the library.

```{r eval=TRUE, warning=FALSE, message=FALSE}
source("setup.R")
```

### Get global climate variables (19 in total) from the [WorldClim Bioclimatic Database](https://www.worldclim.org/data/bioclim.html)

This downloads the 19 climatic variables available on the WorldClim database to your PC, and loads them into R as a SpatRaster object. The ``geodata::worldclim_global()`` code is commented out because the data has already been downloaded for you, and is in the ``data/`` folder. If you were downloading this for the first time, you would need to download from the WorldClim database. This data was downloaded at a 2.5 minute degree resolution. 

```{r eval=FALSE}

# create a data folder in your current working directory

# dir.create("data/environmental_layers/current", recursive = TRUE, showWarnings = FALSE)

# IF THIS IS THE FIRST TIME, DOWNLOAD THE DATA FROM GBIF:
# wc_current = geodata::worldclim_global(
#     var = "bio",
#     res = 2.5,      # Minute degree resolution of raster layers
#     path = here::here("./data/environmental_layers/current/"),
#     version = "2.1"
# )

# IF NOT THE FIRST TIME, load the WORLDCLIM rasters layers already downloaded 
pred_clim = terra::rast( list.files(
  here::here("data/environmental_layers/current/wc2.1_2.5m/") ,
  full.names = TRUE,
  pattern = '.tif'
))  

pred_clim

# set the CRS (coordinate reference system) projection for the current climate layers
terra::crs(pred_clim) = "epsg:4326"
terra::crs(pred_clim, describe = T)

# plot the first variable, annual mean temperature:
terra::plot(pred_clim$wc2.1_2.5m_bio_1)

```

Why do you think setting a consistent CRS projection is important?

We now have all 19 climate variables stored as **pred_clim**. Perhaps we want an additional variable, like elevation. Let's add that to our climate variables as well.

### Get elevation data

Here, we download elevation data in the same way as the climate data above:

```{r eval=FALSE}

# dir.create("data/environmental_layers/elevation/", recursive = TRUE, showWarnings = FALSE)

# pred_altitude = geodata::elevation_global(res = 2.5, path = here::here("data/environmental_layers/elevation/"))

# if already downloaded, read in
pred_elevation = terra::rast(x = here::here(
  "data/environmental_layers/elevation/wc2.1_2.5m/wc2.1_2.5m_elev.tif")
)

# Set the CRS
terra::crs(pred_elevation) = "epsg:4326"
terra::crs(pred_elevation, describe = T)

# plot
terra::plot(pred_elevation)

# check that the CRS for climate and elevation are the same
terra::res(pred_clim)
terra::res(pred_elevation)

```

### Combine climate and elevation data into one raster object

🟢 Create a list called ``raster_list``, into which you add the ``pred_clim`` and ``pred_elevation`` objects

```{r eval=FALSE}

raster_list = # CREATE LIST

# this turns the objects in raster_list into a raster object
predictors = terra::rast(raster_list) 

# Check the variable contains all 20 variables 
terra::nlyr(predictors)

predictors

```

### Download GPS records from GBIF

🟢 Download all GBIF records for **Diaphorina citri** and:

- Keep only the ``species``, ``lat``, and ``lon`` columns
- Save the data frame as an RDS file
- There is a CSV file in the ``data/gps/`` folder called ``Diaphorina_citri.csv``. Read this file into R, and select only the ``species``, ``lat``, and ``lon`` columns. Rename these columns like this: ``dplyr::select(species, lat = decimalLatitude, lon = decimalLongitude)``
- Create a single data frame that contains both the GBIF download data, and the CSV data that you have just read in. Call this data frame ``sp_gps`` 
- Select only the ``species``, ``lon``, and ``lat`` columns, and save that as a new object called ``sp_data``
- Keep only distinct GPS coordinates

```{r eval=FALSE}

# Download all available Diaphorina citri records from GBIF here

# keep just the columns with species name, latitude, and longitude values

# save as RDS

# Read in the CSV file from here: data/gps/Diaphorina_citri.csv

# combine everything
sp_gps = # BIND YOUR DATASETS HERE

# Keep only the species, lon, and lat columns
sp_data = sp_gps %>%
  # SELECT COLUMNS HERE

# remove duplicate GPS records
sp_data = sp_data %>%
  # KEEP DISTINCT LAT AND LON RECORDS

```

Now we need to filter the data further so that we keep only one GPS record per grid cell:

```{r eval=FALSE}

# Convert one of our environmental pred_clim into a raster layer 
r = raster::raster(pred_clim[[1]])

# Extract longitude and latitude into a dataframe
xy = sp_data %>%
  dplyr::select(lon, lat) %>%
  as.data.frame(.)
head(xy)

# Retain only 1 GPS record per grid cell
set.seed(2012)
sp_data_thin = dismo::gridSample(
  xy = xy,     # Data.frame containing lon/lat columns only 
  r = r ,      # Environmental raster (must be a raster, not spatRast object)
  n = 1        # Number of records to keep per cell
)

# let's see how many GPS points we have left in our thinned dataset
nrow(sp_data)
nrow(sp_data_thin)

```

🟢 Compare the number of rows between ``sp_data`` and ``sp_data_thin`` -> How many records were dropped when we extracted one GPS record per grid cell?

Going forward, we'll use ``sp_data_thin`` as our data set of GPS points.

🟢 Fill in the missing bits below to get a map of the world, with our GPS points overlaid onto it. The conditional ``ifelse`` statements in the ``geom_point(aes())`` section give different colours and shapes to the points in the native vs invaded range.

Now save the map as an RDS and PNG file.

```{r eval=FALSE}

# Get a world map
world_map = rnaturalearth::ne_countries(
  scale = "medium", 
  returnclass = "sf"
) 

# Plot GPS points on world map to check our locality data is correct 
global_distr = ggplot() +
  # Add raster layer of world map 
  geom_sf(data = YOURWORLDMAP, alpha = 0.5) +
  # Add GPS points 
  geom_point(
    data = YOURDATA, 
    size = 0.5,
    aes(
      x = lon, 
      y = lat,
      color = ifelse(lon > 60, "red", "black"),
      shape = ifelse(lon > 60, "triangle", "circle")
    )
  )  +
  scale_colour_manual(values = c("red", "black")) +
  # Set world map CRS 
  coord_sf(
    crs = 4326,
    ylim = c(-58, 90),  # Clip below -52 degrees latitude
    expand = FALSE
  ) + 
  xlab("Longitude") + 
  ylab("Latitude")

global_distr

# save the global map 
ggsave(SAVE HERE)

```

🟢 Using your ``sp_data_thin`` data frame, create another data frame called ``species_native``, where you filter the original so that it only keeps records where the longitude is **greater than** 60 degrees East. This object will contain all records from the psyllid's native range in Asia. Do a quick check to see how many records there are in the native range data frame.


```{r eval=FALSE}

species_native = sp_data_thin %>%
  # FILTER LONGITUDE HERE

# CHECK HOW MANY ROWS species_native HAS HERE

```

🟢 Now do the same for the invaded range, where you call the data frame ``species_invaded``. Filter the data so that you keep all records where longitude is **less than** 60 degrees East. Check how many records there are in this invaded range data frame.

```{r eval=FALSE}

# Filter GPS records from the invaded range only

# get just the invaded range (remove all native range GPS coordinates)
species_invaded = sp_data_thin %>%
  # FILTER LONGITUDE HERE
  
# CHECK HOW MANY ROWS species_invaded HAS HERE

```

The code block below just crops all our climate variable data to the size of the native and invaded ranges, so that we only have to work with the data we need (otherwise we retain all the climate data for the whole world, which is unnecessarily clumsy).

```{r eval=FALSE}
species_native_extent = sf::st_as_sf(species_native, coords = c("lon", "lat"), crs = 4326)
species_native_extent = sf::st_bbox(species_native_extent)

# crop the predictors
pred_clim_native_range = raster::crop(predictors, species_native_extent)
plot(pred_clim_native_range$wc2.1_2.5m_bio_1)

species_invaded_extent = sf::st_as_sf(species_invaded, coords = c("lon", "lat"), crs = 4326)
species_invaded_extent = sf::st_bbox(species_invaded_extent)

pred_clim_invaded_range = raster::crop(predictors, species_invaded_extent)
plot(pred_clim_invaded_range$wc2.1_2.5m_bio_1)

```

Here, we can have a look at how many GPS records fall within certain areas, if we provide the bounding boxes (lat and lon) for those regions:

```{r eval=FALSE}

# How many points in the USA?

species_usa = sp_data_thin %>%
  dplyr::filter(lon > -127, lon< -65, lat > -5, lat < 50)

print(paste("There are", nrow(species_usa), "GPS points in the USA.") )

# How many points in Brazil?

species_braz = sp_data_thin %>%
  dplyr::filter(lon > -73.99, lon < -34.80, lat > -33.75, lat < 5.27)

print(paste("There are", nrow(species_braz), "GPS points in Brazil.") )

# How many points in Africa?

species_afr = sp_data_thin %>%
  dplyr::filter(lon > -26, lon < 55, lat > -48, lat < 40)

print(paste("There are", nrow(species_afr), "GPS points in Africa.") )

```

### Multicollinearity: which variables do we choose?

We currently have 20 variables: 19 climate and 1 elevation. There will be some variables that are correlated with each other (e.g. low precipitation might be strongly correlated with high temperature). In these cases, including all correlated variables can be redundant, and it can become difficult to tease apart the more important ones that are actually affecting the model the most. Here, we run a correlation test, and select the variables we decide are the most important for this particular case.

```{r eval=FALSE}

# Extract the climate and elevation data at each presence GPS record for ACP
clim_sp = terra::extract(
  x = predictors,  # SpatRast containing all 19 WORLDCLIM layers + elevation
  y = sp_data_thin # data.frame containing GPS of study taxon (lon, lat)
) %>%
  dplyr::select(-ID) %>% # drop the ID column
  na.omit() # remove any NA values

# create a copy of clim_sp, so that we can run a correlation test on it without altering the original
clim_sp_corr = clim_sp

# change the variable names to a shorter simpler version
new_colnames = gsub(".*_(\\d+)$", "\\1", names(clim_sp_corr))
names(clim_sp_corr) = new_colnames

# order the variables from 1 to 19, rather than 1, 10, 11, etc.
clim_sp_corr = clim_sp_corr[, gtools::mixedsort(names(clim_sp_corr))]

# change the name of the elevation column to make it simpler
clim_sp_corr = clim_sp_corr %>% rename(elevation = wc2.1_2.5m_elev)

```

🟢 Run a Spearman's correlation test on ``clim_sp_corr`` using the ``cor()`` function.

```{r eval=FALSE}

cor_mat = # RUN CORRELATION TEST HERE

```

Let's have a look at the results:

```{r, eval=FALSE}
# have a look at a correlation matrix plot
cor_mat[cor_mat == 1 ] = 0
corrplot::corrplot(cor_mat, type = "upper", order = "original",
         tl.col = "black", tl.srt = 0)

# identify strongly correlated variables
# 0.7 is a standard threshold in the literature, but you can alter this value

CORR.VAL = 0.7
cor_mat[cor_mat < CORR.VAL & cor_mat > -CORR.VAL] = 0

# create a correlation newtork
network = igraph::graph_from_adjacency_matrix(cor_mat, weighted=T, mode="undirected", diag=F)
plot(network, vertex.color = "lightgreen")

```

The connection lines in the network plot show the variables that are strongly correlated, and so we need to select the best ones to continue with. This can be subjective!

Here, for example, bio 1 is correlated with bio 6, 8, and 11. If we choose bio 1, we shouldn't include these other three in our final list. Let's choose these:

**bio 1, bio 2, bio 5, bio 7, bio 9, bio 12, bio 15, bio 19, and elevation.**

As a reminder, these are the bio codes:

**BIO1** = Annual Mean Temperature  
**BIO2** = Mean Diurnal Range (Mean of monthly (max temp - min temp))  
**BIO3** = Isothermality (BIO2/BIO7) (×100)  
**BIO4** = Temperature Seasonality (standard deviation ×100)  
**BIO5** = Max Temperature of Warmest Month  
**BIO6** = Min Temperature of Coldest Month  
**BIO7** = Temperature Annual Range (BIO5-BIO6)  
**BIO8** = Mean Temperature of Wettest Quarter  
**BIO9** = Mean Temperature of Driest Quarter  
**BIO10** = Mean Temperature of Warmest Quarter  
**BIO11** = Mean Temperature of Coldest Quarter  
**BIO12** = Annual Precipitation  
**BIO13** = Precipitation of Wettest Month  
**BIO14** = Precipitation of Driest Month  
**BIO15** = Precipitation Seasonality (Coefficient of Variation)  
**BIO16** = Precipitation of Wettest Quarter  
**BIO17** = Precipitation of Driest Quarter  
**BIO18** = Precipitation of Warmest Quarter  
**BIO19** = Precipitation of Coldest Quarter  


```{r eval=FALSE}

# specify the variables we want to keep
vars_reduced_r2 = c(
  "wc2.1_2.5m_bio_1",
  "wc2.1_2.5m_bio_2",
  "wc2.1_2.5m_bio_5",
  "wc2.1_2.5m_bio_7",
  "wc2.1_2.5m_bio_9",
  "wc2.1_2.5m_bio_12",
  "wc2.1_2.5m_bio_15",
  "wc2.1_2.5m_bio_19",
  "wc2.1_2.5m_elev"
)

# reduce our original full set of predictor variables to the ones specified above
reduced_preds = terra::subset(x = predictors, subset = vars_reduced_r2)

# extract the data for each GPS record
clim_sp_reduced = terra::extract(
  x = reduced_preds,         
  y = sp_data_thin              
) %>%
  dplyr::select(-ID) %>% # drop the ID column
  na.omit() # remove any NA values

# a common approach is to run a follow-up test, using a Variance Inflation Factor method (VIF)
# this is another method of assessing multicollinearity between variables, where a threshold value of 5
# indicates moderate collinearity
usdm::vifstep(clim_sp_reduced, th = 5)

```

The VIF test suggests that we remove bio 1 and bio 7. Since bio 1 is quite important, we'll just remove bio 7.

```{r eval=FALSE}

vars_reduced_r2_VIF = c(
  "wc2.1_2.5m_bio_1",
  "wc2.1_2.5m_bio_2",
  "wc2.1_2.5m_bio_5",
  "wc2.1_2.5m_bio_9",
  "wc2.1_2.5m_bio_12",
  "wc2.1_2.5m_bio_15",
  "wc2.1_2.5m_bio_19",
  "wc2.1_2.5m_elev"
)

# reduce our original full set of predictor variables to the ones specified above
reduced_preds = terra::subset(x = predictors, subset = vars_reduced_r2_VIF)

# write this raster as a tif file
terra::writeRaster(reduced_preds, "reduced_preds.tif")

```

### Presence Points

Now that we have the predictor variables we want, we need to set up a data frame that contains the climate variable data at each recorded presence GPS point. 

```{r eval=FALSE}

# if not still in the R environment, read in our reduced predictor SpatRaster (tif file), and GPS presence points
reduced_preds = terra::rast("data/reduced_preds.tif")
sp_data_thin = readRDS("data/sp_data_thin.rds")

# extract all the data at the GPS points for ACP
speciesEnv = base::data.frame( raster::extract(reduced_preds, 
                cbind(sp_data_thin$lon, sp_data_thin$lat)) )

# add the lat and lon columns to the data
speciesWd = cbind(sp_data_thin, speciesEnv)

# clean the data frame, and add a column with the species name for ACP
speciesWd = as.data.frame(speciesWd) %>%
    dplyr::select(
      decimalLatitude = lat,
      decimalLongitude = lon,
      dplyr::all_of(names(reduced_preds))
    ) %>%
    dplyr::mutate(species = "Diaphorina citri") %>%
    dplyr::select(species, everything()
    ) %>%
  na.omit()

# Reproject the CRS 
coordinates(speciesWd) =  ~ decimalLongitude + decimalLatitude
crs_wgs84 = CRS(SRS_string = "EPSG:4326")
slot(speciesWd, "proj4string") = crs_wgs84

speciesWd = as.data.frame(speciesWd)
head(speciesWd)

# Let's call this presPoints for clarity
presPoints = speciesWd  

# save to file
saveRDS(object = presPoints, file = "data/presPoints.rds")

```

### Background Points (Pseudo-absences)

We need to generate background GPS points where we think we will be unlikely to record ACP. This is quite a debated topic in SDM work, as these pseudo-absence points are generated based on statistical approaches, rather than actual verified data. Nevertheless, a good approach is to select points from regions that are at least somewhat similar to presence areas in terms of climate type. Here, we'll use Köppen-Geiger climate data to select similar eco-regions to randomly sample points from.

```{r eval=FALSE}

# read in the koppen-geiger (KG) data
kg_layer = sf::st_read("koppen_geiger")

# Reproject KG layer
geo_proj = "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
kg_layer = sf::st_transform(kg_layer, geo_proj)

# Coerce GPS records into SpatialPointsDataFrame
records_spatial = sp::SpatialPointsDataFrame(
  coords = cbind(sp_data_thin$lon, sp_data_thin$lat),
  data = sp_data_thin,
  proj4string = CRS('+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0')
)

# Select KG ecoregions in which there is at least one GPS record
kg_sp = as(kg_layer, "Spatial")
kg_contain = kg_sp[records_spatial, ]

# Define background area by masking predictor layers to just the KG zones with at 
# least 1 GPS record
# Convert the KG zones containing GPS records back into an 'sf' object
crs_wgs84 = CRS(SRS_string = "EPSG:4326")
kg_contain = sf::st_as_sf(kg_contain)
kg_contain = sf::st_set_crs(kg_contain, crs_wgs84)

bg_area = terra::mask(reduced_preds, kg_contain)  

terra::writeRaster(bg_area, "data/bg_area.tif")

```

```{r eval=FALSE}

presPoints = readRDS("data/presPoints.rds")
bg_area = terra::rast("data/bg_area.tif")

# Plot to check the mask worked -> this is the area where we will pick background points from
terra::plot(bg_area$wc2.1_2.5m_bio_1)

```

Now that we have an area from which to select background points, we can go ahead and randomly select 10,000 points (without replacement). Then, as before, we will extract our predictor variable data at those points and create a SpatRaster .tif file.

```{r eval=FALSE}

# setting a seed makes this process repeatable, if you re-run this again
set.seed(2023)
  
bg_points = terra::spatSample(
    x = bg_area,        # Raster of background area to sample points from 
    size = 10000,       # How many background points do we want?
    method = "random",  # Random points
    replace = FALSE,    # Sample without replacement
    na.rm = TRUE,       # Remove background points that have NA climate data
    as.df = TRUE,       # Return background points as data.frame object
    xy = TRUE           # Return lat/lon values for each background point
  ) %>%
dplyr::rename(lon = x, lat = y)
  
head(bg_points)

terra::plot(kg_contain, col="khaki")
points(bg_points$lon, 
       bg_points$lat, 
       pch = 21, 
       bg = 'black', 
       cex = 0.2)
points(sp_data_thin$lon, 
       sp_data_thin$lat, 
       pch = 21, 
       bg = 'red', 
       cex = 1)

```

If you want to make sure that no background points are selected at the same position as a presence point, you can use the ``predicts::backgroundSample()`` function shown below. 

```{r eval=FALSE}
bg_points2 = predicts::backgroundSample(
  mask = bg_area,
  n = 10000,
  p = sp_data_thin,
  excludep = TRUE
)

head(bg_points2)
``` 

Alternatively, you could take your ``bg_points`` from before, compare it to your presence points in ``sp_data_thin``, and remove any lon-lat pairs that are identical:

```{r eval=F}
# Assume df1 and df2 have columns "lon" and "lat"
common_points = dplyr::inner_join(bg_points, sp_data_thin, by = c("lon", "lat"))

# View how many matches were found
nrow(common_points)
```

Here, none are identical, so we can proceed.

```{r eval=F}

# clean the data frame, and add a column with the species name for ACP
bg_pointsWd = as.data.frame(bg_points) %>%
    dplyr::select(
      decimalLatitude = lat,
      decimalLongitude = lon,
      dplyr::all_of(names(reduced_preds))
    ) %>%
    dplyr::mutate(species = "Diaphorina citri") %>%
    dplyr::select(species, everything() ) 

# Reproject CRS 
coordinates(bg_pointsWd) =  ~ decimalLongitude + decimalLatitude
crs_wgs84 = CRS(SRS_string = "EPSG:4326")
slot(bg_pointsWd, "proj4string") = crs_wgs84

bg_pointsWd = as.data.frame(bg_pointsWd)
head(bg_pointsWd)  

# Let's call this backPoints for clarity
backPoints = bg_pointsWd 

# save to file
saveRDS(object = backPoints, file = "data/backPoints.rds")

```

As an aside, let's add some buffering around presence points, and remove any background points that fall within that buffer. Let's try a 300 km buffer zone around each presence point:

```{r eval=F}

world_map = rnaturalearth::ne_countries(
  scale = "medium", 
  returnclass = "sf"
) 

# Convert data frames to sf objects
pres_sf = sf::st_as_sf(sp_data_thin, coords = c("lon", "lat"), crs = 4326)
bg_sf = sf::st_as_sf(bg_points, coords = c("lon", "lat"), crs = 4326)

# Transform to projected CRS (for distance-based buffering in meters)
pres_sf_proj = sf::st_transform(pres_sf, crs = 3395)
bg_sf_proj   = sf::st_transform(bg_sf, crs = 3395)

# Create buffer around presence points (in meters)
# let's apply a 300 km buffer zone here. You can experiment with other values
pres_buffer = sf::st_buffer(pres_sf_proj, dist = 300000)

# Transform buffers back to geographic coordinates for plotting
pres_buffer_latlon = sf::st_transform(pres_buffer, crs = 4326)

# Combine buffers into a single geometry
pres_union = st_union(pres_buffer)

# Filter out background points that fall within the buffer
bg_filtered = bg_sf_proj[!st_intersects(bg_sf_proj, pres_union, sparse = FALSE), ]

# Transform back to geographic coordinates
bg_filtered_latlon = st_transform(bg_filtered, crs = 4326)

########################
# before buffering
########################

ggplot() +
  geom_sf(data = pres_buffer_latlon, fill = "royalblue", alpha = 0.3, color = NA) +  # translucent buffer
  geom_sf(data = world_map, alpha=0.5) +
  geom_sf(data = pres_sf, color = "black", size = 2) +  # presence points
  geom_sf(data = bg_sf, color = "darkred", size = 0.5) +
  labs(title = "Presence Points with 300 km Buffers")

########################
# after buffering
########################

buffer.map = ggplot() +
  geom_sf(data = pres_buffer_latlon, fill = "royalblue", alpha = 0.3, color = NA) +  # translucent buffer
  geom_sf(data = world_map, alpha=0.5) +
  geom_sf(data = pres_sf, color = "black", size = 2) +  # presence points
  geom_sf(data = bg_filtered_latlon, color = "darkred", size = 0.5) +
  labs(title = "Presence Points with 300 km Buffers")

saveRDS(buffer.map, "data/buffer.map.rds")

# Convert back to data frame 
backPoints.buffer = bg_filtered_latlon %>%
 dplyr::mutate(lon = st_coordinates(.)[,1],
         lat = st_coordinates(.)[,2]) %>%
  sf::st_drop_geometry()

# save to file
saveRDS(object = backPoints.buffer, file = "data/backPoints.buffer.rds")

```

With the buffering, you should get a map like this:

```{r eval=TRUE, echo=FALSE}

buffer.map = readRDS("data/buffer.map.rds")
buffer.map

```

### Setup for Model Tuning

We have our presence and background GPS points for ACP, and now we need to find the best model parameters for running MaxEnt. This involves finding the optimal Feature Class (FC) and Regularisation Multiplier (RM) values, which will ensure that the predictions made are as accurate as they can be, given the input data.

```{r eval=FALSE}

# this bit of code creates folders in your working directory (WD), into which you can save output text files and figures
# it won't overwrite a folder if it already exists

if(!dir.exists("tuning")){
  dir.create("tuning")
}

# let's read in our presence and background points from our WD
presPoints = readRDS("data/presPoints.rds")
backPoints = readRDS("data/backPoints.rds")
# and our predictors again
reduced_preds = terra::rast("data/reduced_preds.tif")

# we only need the lat and lon columns for the model tuning, so pull those out
PRES.GPS = presPoints %>% dplyr::select(lon = decimalLongitude, lat = decimalLatitude)
BACK.GPS = backPoints %>% dplyr::select(lon = decimalLongitude, lat = decimalLatitude)

# combine the presence and background data so that you can get the extent of both
PRES_and_BACK = rbind(PRES.GPS, BACK.GPS)

PRES_and_BACK_extent = sf::st_as_sf(PRES_and_BACK, coords = c("lon", "lat"), crs = 4326)
PRES_and_BACK_extent = sf::st_bbox(PRES_and_BACK_extent)

# crop the predictor data so that it covers just the necessary extent 
# of the occurrence + background data
# this just makes the data a bit smaller and more manageable, as we don't need the full predictor extent
predictor_data = raster::crop(reduced_preds, PRES_and_BACK_extent) %>%
  raster::stack()

terra::plot(predictor_data$wc2.1_2.5m_bio_1)

```

### Model Tuning

Now we are ready to start the model tuning process. This can take minutes to hours, depending on the size of the data, and how many FC and RM values you want to test. Sometimes this can cause your R session to crash, if it is too memory intensive for your computer. Setting the java.parameters option could help, otherwise you may need to restart R.

Have a look at how spatial folds are applied by ``ENMevaluate`` when tuning model paramaters.
Presence and background points are divided into four major geographic blocks. Each fold uses one block as test data and the remaining three blocks as training data ENMeval performs cross-validation across these four folds, training a model on three blocks and testing on the held-out block, repeating for each block. Performance metrics (e.g., test AUC) are averaged across folds.

There are other methods available, such as the checkerboard approach. See the ENMeval manual: https://jamiemkass.github.io/ENMeval/articles/ENMeval-2.0-vignette.html

```{r eval=F}

block = ENMeval::get.block(PRES.GPS, BACK.GPS, orientation = "lat_lon")
table(block$occs.grp)
ENMeval::evalplot.grps(pts = BACK.GPS, pts.grp = block$bg.grp, envs = predictor_data) + 
  ggplot2::ggtitle("Spatial block partitions: background")
ENMeval::evalplot.grps(pts = PRES.GPS, pts.grp = block$occs.grp, envs = predictor_data) + 
  ggplot2::ggtitle("Spatial block partitions: presences")

occs.z = cbind(PRES.GPS, terra::extract(predictor_data, PRES.GPS, ID = FALSE))
bg.z = cbind(BACK.GPS, terra::extract(predictor_data, BACK.GPS, ID = FALSE))
evalplot.envSim.hist(occs.z = occs.z, bg.z = bg.z, occs.grp = block$occs.grp, 
                     bg.grp = block$bg.grp, ref.data = "occs")
```

This code block shows you how you would run model tuning from scratch. You don't need to run this block - the model has already been made for you.

```{r eval=FALSE}

# helps with memory allocation for Java
options(java.parameters = "-Xmx16000m")

list_settings = list(
  fc = c("L","Q","H","LQH"), 
  rm = c(1:4) # RM values from 1 to 4 should be sufficient
)

set.seed(2023)
dismo::maxent()

# Run the tuning analysis

tuning_results =
  ENMeval::ENMevaluate(
    occs = PRES.GPS,
    envs = predictor_data,
    bg = BACK.GPS,
    tune.args = list_settings,
    partitions = "block",
    algorithm = "maxent.jar",
    doClamp = FALSE
  )

res = ENMeval::eval.results(tuning_results)
res = na.omit(res)

# save to file
write.csv(res, file = "tuning/model_tuning_enmeval_full.csv",
          quote = FALSE, row.names = FALSE)

# extract key metrics
subset_res = res %>%
  dplyr::select(rm, fc, auc.val.avg, auc.diff.avg, cbi.val.avg, or.10p.avg, AICc)

# alter the format
df_long = subset_res %>%
  tidyr::gather(key = "variable", value = "value", -rm, -fc)
df_long$rm = as.factor(df_long$rm)

# save to file
write.csv(df_long, file = "tuning/model_tuning_enmeval.csv",
          quote = FALSE, row.names = FALSE)

``` 

```{r eval=TRUE}

# I have already run this tuning analysis, and so we can read the results in to proceed

df_long = read.csv("tuning/model_tuning_enmeval.csv")

tuning_results_ggplot = ggplot(df_long, aes(x = rm, y = value, color = fc, group = fc)) +
  geom_line() +
  geom_point(aes(shape = fc)) +
  scale_shape_manual(values = c(15, 16, 17, 5)) +
  facet_wrap(~variable, scales = "free_y", nrow = 4) +
  scale_colour_manual(values = c("black", "darkred", "royalblue", "forestgreen")) +
  labs(x = "Regularisation multiplier", y = "Value") +
  theme_classic() ;tuning_results_ggplot

# ggsave("figs/tuning_results.png", tuning_results_ggplot, width = 9, height = 7)
```

Find the best models, optimising different metrics.

```{r eval=FALSE}

res = read.csv("tuning/model_tuning_enmeval_full.csv")

# Select the model settings (RM and FC) that optimised AICc (delta AICc == 0)
best_delta_aicc = res %>% 
  dplyr::filter(delta.AICc == 0) ;best_delta_aicc
best_delta_aicc_output = best_delta_aicc %>% t(.)
write.table(best_delta_aicc_output, "tuning/LQH1_model.txt", quote = FALSE)

# select the model that optimised AUC (highest AUC value)
best_auc = res %>% 
  dplyr::filter(auc.val.avg == max(auc.val.avg)) ;best_auc
best_auc_test_output = best_auc %>% t(.)
write.table(best_auc_test_output, "tuning/H3_model.txt", quote = FALSE)

# select the model that optimised CBI (highest CBI value)
best_cbi = res %>% 
  dplyr::filter(cbi.val.avg == max(cbi.val.avg)) ;best_cbi
best_cbi_output = best_cbi %>% t(.)
write.table(best_cbi_output, "tuning/H1_model.txt", quote = FALSE)

# select the model that optimised the 10% omission rate (lowest or.10p value)
best_or.10p.avg = res %>% 
  dplyr::filter(or.10p.avg == min(or.10p.avg)) ;best_or.10p.avg
best_or.10p.avg_output = best_or.10p.avg %>% t(.)
write.table(best_or.10p.avg_output, "tuning/H7_model.txt", quote = FALSE)

# default model output
default_mod_results = res %>% 
  dplyr::filter(tune.args == "fc.LQH_rm.1") 
default_mod_results = default_mod_results %>% t(.)
write.table(default_mod_results, "tuning/best_model_default.txt", quote = FALSE)

```

Area under the curve (AUC) values are commonly used to report model accuracy. Here are the AUC values for each of the outputs above, optimising for different metrics:

| Metric  | Model | Value |
|:-------:|:-----:|:-----:|
| AICc    | LQH1  | 0.84  |
| AUC     | H3    | 0.85  |
| CBI     | H1    | 0.84  |
| OR10    | H7    | 0.84  |
| Default | LQH1  | 0.84  |

The model optimising for AUC (FC = H, RM = 3) is the best, with an AUC value of 85% accuracy.

There are auc.train, auc.diff.avg, auc.diff.sd, auc.val.avg, and auc.val.sd values. We're interested in the auc.train (accuracy on the training data) and auc.val.avg (accuracy on the test data) values. The auc.val.sd value of 0.03 shows the variation on the four-fold cross validation approach we took earlier in ``ENMeval``.

### Principal Component Analysis - explore predictor variables

Let's have a look at the contribution of each variable to the model.

```{r eval=FALSE}

presPoints = readRDS("data/presPoints.rds")
climate.df = dplyr::select(presPoints, !c(species, decimalLatitude, decimalLongitude))
head(climate.df)

# Create new column names
new_names = c(paste0("bio_", c("1", "2", "5", "9", "12", "15", "19")), "elev")

# Set the new column names
colnames(climate.df) = new_names
head(climate.df)

```

🟢 Run a PCA on the ``climate.df`` data frame using the ``prcomp()`` function, and then access the loadings using the ``rotation`` attribute of ``pca_res`` (hint: use the dollar sign to access attributes -> ``$``).

```{r eval = FALSE}

pca_res = # RUN PCA HERE
summary(pca_res)

loadings = # ACCESS LOADINGS
loadings

PCA = factoextra::fviz_pca_var(pca_res,
                  col.var = "contrib", # Color by contributions to the PC
                  gradient.cols = c("blue", "gold", "red"),
                  repel = TRUE) + theme_classic(); PCA

# Contributions of variables to PC1
a = factoextra::fviz_contrib(pca_res, choice = "var", axes = 1)
# Contributions of variables to PC2
b = factoextra::fviz_contrib(pca_res, choice = "var", axes = 2)
pca_contribs = gridExtra::grid.arrange(a, b, ncol=2, top='Contribution of the variables to the first two PCs')

#ggsave(file = "figs/pca_contributions.png", pca_contribs, dpi = 350, height = 5, width = 10)
#ggsave(file = "figs/pca.png", PCA, dpi = 350, height = 5, width = 6)

```

Let's plot the PCA again, but now let's have a look at whether there is any difference between the climate in the native versus the invaded range. First, we'll create a new data frame from our presence points, and add a grouping column called ``range``. We can use an ``ifelse`` statement here to assign the word "native" or "nonnative" to each GPS record depending on its longitude:

```{r eval=FALSE}

presPoints.new = presPoints %>%
dplyr::mutate(range = dplyr::if_else(decimalLongitude > 60, "native", "nonnative"))

head(presPoints.new)

head(climate.df)

# extract all PC1 and PC2 values
pca_res_df = pca_res$x %>%
  as.data.frame() %>%
  # keep only PC1 and PC2
  dplyr::select(c(PC1, PC2))

# add the range grouping column
pca_res_df$range = presPoints.new$range

pca_plot = pca_res_df %>%
  ggplot(aes(PC1, PC2, color = range)) + 
  geom_point(size = 1.5, alpha = 0.5) +
  scale_color_manual(
    values = c(nonnative = "darkred", native = "forestgreen")
  ) +
  xlab("PC1 (32%)") +
  ylab("PC2 (31%)") +
  labs(color = "Range",
       title = "ACP niche conservatism") +
  # add lines through the origin 
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  theme_classic()

pca_plot

# another approach, adding ellipses
factoextra::fviz_pca_ind(pca_res, geom = "point",
                         col.ind = presPoints.new$range,
                         palette = c("forestgreen",  "darkred"),
                         legend.title = "Range", alpha.ind = 0.5,
                         addEllipses = TRUE) +
                         scale_shape_manual(values=c(19,17))

# run t-tests to see whether there is a difference between native vs non-native points on PC1, and PC2
t.test(PC1 ~ range, data = pca_res_df)
t.test(PC2 ~ range, data = pca_res_df)

# looks like there is a difference on PC1, but not PC2

# check the distribution of the data
hist(pca_res_df$PC1)
hist(pca_res_df$PC2)

```

Run a Schoener's D test to check niche overlap:

```{r eval=FALSE}
pca_native = pca_res_df %>%
  dplyr::filter(range == "native") %>%
  dplyr::select(PC1, PC2)

pca_nonnative = pca_res_df %>%
  dplyr::filter(range == "nonnative") %>%
  dplyr::select(PC1, PC2)

all_pca = pca_res_df %>%
  dplyr::select(PC1, PC2)

grid_native = ecospat::ecospat.grid.clim.dyn(glob = all_pca, glob1 = pca_native, sp = pca_native, R = 100)
grid_nonnative = ecospat::ecospat.grid.clim.dyn(glob = all_pca, glob1 = pca_nonnative, sp = pca_nonnative, R = 100)

# Schoener's D
schoener_d = ecospat::ecospat.niche.overlap(grid_native, grid_nonnative, cor = FALSE)$D

# Print the overlap value
print(schoener_d)

ecospat::ecospat.plot.niche.dyn(grid_native, grid_nonnative, quant = 0.1, interest = 2, title = "Niche Overlap", name.axis1 = "PC1", name.axis2 = "PC2", col.unf = "black", col.exp = "darkred", col.stab = "lightgreen", colZ1 = "darkgreen", colZ2 = "red", transparency = 60)
```

**Schoener's D interpretation:**

0.0 → No overlap (completely distinct ecological niches)    
0.5 → Moderate overlap (some shared environmental space)    
1.0 → Complete overlap (both groups use the same niche)   

Let's make some box-plots, showing each bioclimate variable (and elevation) across the native vs non-native range:

```{r eval=FALSE}

presPoints_long = presPoints.new %>%
  tidyr::pivot_longer(cols = starts_with("wc2.1_2.5m_bio") | "wc2.1_2.5m_elev",  # Include elevation
               names_to = "bio_variable", 
               values_to = "value")

new_names = c(paste0("bio_", c("1", "2", "5", "9", "12", "15", "19")), "elev")

# Reorder and rename the bio_variable levels
presPoints_long$bio_variable = factor(presPoints_long$bio_variable,
                                       levels = c("wc2.1_2.5m_bio_1", "wc2.1_2.5m_bio_2", "wc2.1_2.5m_bio_5", 
                                                  "wc2.1_2.5m_bio_9", "wc2.1_2.5m_bio_12", "wc2.1_2.5m_bio_15", 
                                                  "wc2.1_2.5m_bio_19", "wc2.1_2.5m_elev"),
                                       labels = new_names)

# Plot boxplots for native vs non-native groups, faceted by bio variables and including elevation
range.boxes = ggplot(presPoints_long, aes(x = range, y = value, fill = range)) +
  geom_boxplot(alpha = 0.5) +
  scale_fill_manual(values = c(nonnative = "darkred", native = "forestgreen")) +
  facet_wrap(~ bio_variable, scales = "free_y") +  # Facet by bio variable with independent y-scales
  labs(x = "Range (Native vs Non-native)", y = "Value", 
       title = "ACP Bioclimatic Variables and Elevation by Range")

range.boxes 

```

## Create the MaxEnt model

We are finally ready to create a MaxEnt model, which we can project onto any geographic area we like. Keep in mind though the distinction between model interpolation and extrapolation, and the reliability of the model considering the range of values that it was trained on.

Our best model was H3, and so we will set ``'betamultiplier=3.0'`` and ``'hinge=true'``.

### Fit the MaxEnt model

```{r eval=FALSE}

presPoints = readRDS("data/presPoints.rds")
backPoints = readRDS("data/backPoints.rds")
reduced_preds = terra::rast("data/reduced_preds.tif")

# bind the presence and absence data together into one data frame
maxent_data = dplyr::bind_rows(presPoints, backPoints) %>% 
  dplyr::select(-c(species, decimalLatitude, decimalLongitude)) 
rownames(maxent_data) = NULL
head(maxent_data)
nrow(maxent_data)

# Create a vector containing 0 (indicating background points) and 1 (indicating presence points)
maxent_vector = c(
  replicate(nrow(presPoints), "1"),
  replicate(nrow(backPoints), "0")
) 

# FIT MAXENT MODEL
maxentModel1 = dismo::maxent(
  x = maxent_data,
  p = maxent_vector,
  path = here::here("models/maxent_H3/"),
  replicates = 10,
  args = c(
    # Insert the optimal RM value here
    'betamultiplier=3.0',
    # Turn these on/off to change FC combinations 
    # - To only use quadratic features, turn all to false except quadratic
    'linear=false',
    'quadratic=false',
    'product=false',
    'threshold=false',
    'hinge=true',
    # Don't change anything from here down 
    'threads=2',
    'doclamp=true',
    'fadebyclamping=true',
    'responsecurves=true',
    'jackknife=true',
    'askoverwrite=false',
    'responsecurves=true',
    'writemess=true',
    'writeplotdata=true',
    'writebackgroundpredictions=true'
  )
)

response_data = dismo::response(maxentModel1, expand = TRUE)

# get the response data for each variable
bio1.response = as.data.frame(dismo::response(maxentModel1, var = "wc2.1_2.5m_bio_1"))  %>%
  dplyr::mutate(name = "bio1") %>% dplyr::rename(value = V1)

bio2.response = as.data.frame(dismo::response(maxentModel1, var = "wc2.1_2.5m_bio_2"))%>%
  dplyr::mutate(name = "bio2") %>% dplyr::rename(value = V1)

bio5.response = as.data.frame(dismo::response(maxentModel1, var = "wc2.1_2.5m_bio_5"))%>%
  dplyr::mutate(name = "bio5") %>% dplyr::rename(value = V1)

bio9.response = as.data.frame(dismo::response(maxentModel1, var = "wc2.1_2.5m_bio_9")) %>%
  dplyr::mutate(name = "bio9") %>% dplyr::rename(value = V1)

bio12.response = as.data.frame(dismo::response(maxentModel1, var = "wc2.1_2.5m_bio_12"))%>%
  dplyr::mutate(name = "bio12") %>% dplyr::rename(value = V1)

bio15.response = as.data.frame(dismo::response(maxentModel1, var = "wc2.1_2.5m_bio_15"))%>%
  dplyr::mutate(name = "bio15") %>% dplyr::rename(value = V1)

bio19.response = as.data.frame(dismo::response(maxentModel1, var = "wc2.1_2.5m_bio_19"))%>%
  dplyr::mutate(name = "bio19") %>% dplyr::rename(value = V1)

elev.response = as.data.frame(dismo::response(maxentModel1, var = "wc2.1_2.5m_elev"))%>%
  dplyr::mutate(name = "elevation") %>% dplyr::rename(value = V1)

# join everything together
bioresponses = rbind(bio1.response, bio2.response, bio5.response, bio9.response, bio12.response, bio15.response,
                     bio19.response, elev.response) %>%
  dplyr::mutate(name = factor(name, levels = c("bio1", "bio2", "bio5", "bio9", "bio12", "bio15", "bio19", "elevation")))

# make a custom ggplot
response.plots = ggplot(bioresponses, aes(x = value, y = p)) +
  geom_line(linewidth = 1, colour = "darkgrey") +
  facet_wrap(~name, scales = "free") +
  labs(x = "Predictor Value", y = "Predicted Response", 
       title = "Maxent Response Curves")

response.plots

```

Examine the model output files, particularly the response curves for each variable, and the importance of each variable to the model (contribution scores). Here, bio 1, 12, and 19 appear to have the greatest influence on the model. The average AUC result is 0.83 (i.e. 83% accuracy on the training data).

### Set up the prediction area

Now that we have created our MaxEnt model, we need to decide where we want to project it, and then create raster layers for that area using our set of predictor variables. Below, we will create raster layers for the African continent:

```{r eval = FALSE}

if(!dir.exists("africa_env_layers")){
  dir.create("africa_env_layers")
}

if(!dir.exists("africa_env_layers/current")){
  dir.create("africa_env_layers/current")
}

# get the raster for Africa
africa_ext = rnaturalearth::ne_countries(scale = "medium", returnclass = "sf") %>%
  dplyr::filter(continent == "Africa")

# read in our predictors again
reduced_preds = terra::rast("data/reduced_preds.tif")

# crop rasters
africa_env_layers = raster::crop(reduced_preds, raster::extent(africa_ext))

# set the CRS
crs(africa_env_layers) = "EPSG:4326"

terra::writeRaster(africa_env_layers, "africa_env_layers/africa_env_layers.tif", overwrite=TRUE)

```

```{r eval=FALSE}

terra::rast("africa_env_layers/africa_env_layers.tif") %>% plot()

```

```{r eval = FALSE}

directory.current = "africa_env_layers/current/"

# Loop over each climate variable and save it with the correct name
# write these rasters to file
for (p in names(reduced_preds)){
  raster::writeRaster(
    africa_env_layers[[which(names(africa_env_layers) %in% p)]],
    filename = paste0(directory.current, p, ".grd"),
    bylayer = TRUE,
    overwrite = TRUE,
    format = "raster",
    bandorder = 'BIL'
  )
}

```

## Project the MaxEnt model

Finally we can do some projecting!

```{r eval=FALSE}

if(!dir.exists("raster_projections")){
  dir.create("raster_projections")
}

africa_env_layers = terra::rast("africa_env_layers/africa_env_layers.tif")

# get the raster for Africa
africa_ext = rnaturalearth::ne_countries(scale = "medium", returnclass = "sf") %>%
  dplyr::filter(continent == "Africa")

africa_ext = sf::st_set_crs(africa_ext, 4326)

# use the model to predict climate suitability for Africa
predict_maxent_africa = terra::predict(maxentModel1, africa_env_layers)
predict_maxent_africa = terra::mask(predict_maxent_africa, africa_ext)

terra::plot(predict_maxent_africa)

terra::writeRaster(predict_maxent_africa, "raster_projections/maxent_projection_africa.tif",
                   overwrite = TRUE)

presPoints = readRDS("data/presPoints.rds")

# filter to get points just in Africa
species_afr = presPoints %>%
  dplyr::filter(decimalLongitude > -26, decimalLatitude < 55, decimalLatitude > -48, decimalLatitude < 40)

africa = ggplot() +
  # Plot MaxEnt prediction raster
  tidyterra::geom_spatraster(
    data = predict_maxent_africa,
    maxcell = 5e+7         # maxcell = Inf
  ) +
  # Control raster colour and legend
 tidyterra::scale_fill_whitebox_c(
    palette = "muted",
    breaks = seq(0, 1, 0.2),
    limits = c(0, 1)
  ) +
  # Plot Africa boundary
  geom_sf(data = africa_ext, fill = NA, color = "black", size = 0.1) +
  # Control axis and legend labels 
  labs(
    x = "Longitude",
    y = "Latitude",
    fill = "P(suitability)"
  ) +
  geom_point(
    data = species_afr,
    size = 1,
    aes(x = decimalLongitude, y = decimalLatitude)
  )  +
  # Crops map to just the geographic extent of Africa
  coord_sf(
    xlim = c(-25, 55),
    ylim = c(-35, 38),
    crs = 4326,
    expand = FALSE
  ) +
  # Create title for the legend
  theme(legend.position = "right") +
  # Add scale bar to bottom-right of map
 ggspatial::annotation_scale(
    location = "bl",          # 'bl' = bottom left
    style = "ticks",
    width_hint = 0.2
  ) +
  # Add north arrow
  ggspatial::annotation_north_arrow(
    location = "bl",
    which_north = "true",
    pad_x = unit(0.175, "in"),
    pad_y = unit(0.3, "in"),
    style = ggspatial::north_arrow_fancy_orienteering
  ) +
  # Change appearance of the legend
  guides(
    fill = guide_colorbar(ticks = FALSE)
  ) ;africa

# save the figure
ggsave("figures/africa_map.png", plot = africa, width = 6, height = 6, dpi = 350)
saveRDS(africa, "figures/africa_map.rds")

```

Save in a format that can be opened in Google Earth as a KML layer. Open this file in Google Earth to have a look at your prediction layer! 

```{r eval=FALSE}

africa_kml = terra::project(predict_maxent_africa, "EPSG:4326")
africa_kml = raster::raster(africa_kml)
raster::KML(africa_kml, "figures/africa", maxpixels=5000000, overwrite = TRUE)

```

Let's write a quick function to plot maps. This saves us the trouble of re-using the same ggplot code over and over again; making the coding process more efficient. Every time we want to plot a raster prediction layer, we can just call this function. Let's call it ``plot.predmap``:

```{r, eval=FALSE}

plot.predmap = function(raster, EXT, title = "", arrow.pos = "bl", scale.pos = "bl"){
  
 MAP = ggplot() +
  # Plot MaxEnt prediction raster
  tidyterra::geom_spatraster(
    data = raster,
    maxcell = 5e+7         # maxcell = Inf
  ) +
  # Control raster colour and legend
 tidyterra::scale_fill_whitebox_c(
    palette = "muted",
    breaks = seq(0, 1, 0.2),
    limits = c(0, 1)
  ) +
  # Plot boundary
  geom_sf(data = EXT, fill = NA, color = "black", size = 0.1) +
  # Control axis and legend labels 
  labs(
    title = title,
    x = "Longitude",
    y = "Latitude",
    fill = "P(suitability)"
  ) +
  # Crops map to just the geographic extent of Africa
  coord_sf(
    xlim = c(raster::extent(EXT)[1],raster::extent(EXT)[2]),
    ylim = c(raster::extent(EXT)[3],raster::extent(EXT)[4]),
    crs = 4326,
    expand = FALSE
  ) +
  # Create title for the legend
  theme(legend.position = "right") +
  # Add scale bar to bottom-right of map
 ggspatial::annotation_scale(
    location = scale.pos,          # 'bl' = bottom left
    style = "ticks",
    width_hint = 0.2
  ) +
  # Add north arrow
  ggspatial::annotation_north_arrow(
    location = arrow.pos,
    which_north = "true",
    pad_x = unit(0.175, "in"),
    pad_y = unit(0.3, "in"),
    style = ggspatial::north_arrow_fancy_orienteering
  ) +
  # Change appearance of the legend
  guides(
    fill = guide_colorbar(ticks = FALSE)
  )
  
  return(MAP)
}

```

A quick example of how one would use the function to plot the prediction for Africa:

``` {r, eval=FALSE}

africa = plot.predmap(raster = predict_maxent_africa, EXT = africa_ext, title = "Africa")

# if we want to crop it just a bit more, so that we don't have a large space at the bottom (because Marion island is included in South Africa), we can add a further ylim change:

africa = africa +
  coord_sf(
    ylim = c(-35,38),
    crs = 4326,
    expand = FALSE
  ) 

africa

```

### Some quick stats: how well did the model perform?

#### How well did the model perform for the African continent?

```{r eval=FALSE}

predict_maxent_africa = terra::rast("raster_projections/maxent_projection_africa.tif")

# convert presence and background points into vector objects
prespoints.vect = terra::vect(
  presPoints,
  geom = c("decimalLongitude", "decimalLatitude"),
  crs = "EPSG:4326"
)

backpoints.vect = terra::vect(
  backPoints,
  geom = c("decimalLongitude", "decimalLatitude"),
  crs = "EPSG:4326"
)
  
##############################################################
# Extract MaxEnt suitability scores at each PRESENCE GPS point
##############################################################
  
maxent_pres_scores = terra::extract(
      x = predict_maxent_africa, # prediction raster of Africa
      y = prespoints.vect,
      xy = TRUE,
      na.rm = TRUE
    ) %>%
    # Clean df
    dplyr::select(
      lat = y,
      lon = x,
      suit_score = 2 # assign the name of the second column ("median") to "suit_score"
    ) %>%
    tidyr::drop_na(suit_score) %>%
    dplyr::mutate(pres = 1)
  
##############################################################
# Extract MaxEnt suitability scores at each BACKGROUND GPS point
##############################################################
  
maxent_background_scores = terra::extract(
      x = predict_maxent_africa,
      y = backpoints.vect,
      xy = TRUE,
      na.rm = TRUE
    ) %>%
    # Clean df
    dplyr::select(
      lat = y,
      lon = x,
      suit_score = 2
    ) %>%
    tidyr::drop_na(suit_score) %>%
    dplyr::mutate(pres = 0)

##############################################################
# Combine dataframes 
##############################################################

clim_data.africa = dplyr::bind_rows(maxent_pres_scores, maxent_background_scores)

mod.africa = glm(pres ~ suit_score, data = clim_data.africa, family = binomial(link = "logit"))
  
print(summary(mod.africa))

# get the beta coefficient
beta_original = coef(mod.africa)["suit_score"]
beta_per_1percent = beta_original * 0.01
odds_ratio_1percent = exp(beta_per_1percent)
cat("Odds ratio for a 1% increase:", odds_ratio_1percent)

# The value of 1.057333 means that the odds of recording the presence of ACP increases by a factor of 1.057333 with a 1% increase in MaxEnt suitability score.

# check model fit
DHARMa::simulateResiduals(fittedModel = mod.africa, plot = TRUE)

car::Anova(mod.africa, test = "LR", type = "II")
# the LR test shows a significant p-value
  
##############################################################
# Plot statistical model  
##############################################################
  
# Extract model predictions 
preds.africa = ggeffects::ggpredict(mod.africa, terms = c("suit_score [0:1 by = 0.01]")) %>%
    as.data.frame() %>%
    dplyr::rename(suit_score = x)
  
clim_data_africa_presences = dplyr::filter(clim_data.africa, pres == 1)
clim_data_africa_absences = dplyr::filter(clim_data.africa, pres == 0)


#################################
# sensitivity
#################################

# find the optimal threshold, based on the model
e.africa = dismo::evaluate(p = presPoints, a = backPoints, model = maxentModel1)

# View various thresholds
thresholds.africa = dismo::threshold(e.africa)
print(thresholds.africa)

# Define a threshold
threshold.africa = 0.35
predicted_presence.africa = ifelse(clim_data.africa$suit_score >= threshold.africa, 1, 0)

# Create confusion matrix
conf_matrix.africa = table(Predicted = predicted_presence.africa, Observed = clim_data.africa$pres)
print(conf_matrix.africa)

# Extract TP and FN
TP.africa = conf_matrix.africa["1", "1"]
FN.africa = conf_matrix.africa["0", "1"]

# Calculate sensitivity
sensitivity.africa = TP.africa / (TP.africa + FN.africa)
cat("Sensitivity:", sensitivity.africa, "\n")

# Plot model predictions 
prob_curve.africa = preds.africa %>%
    ggplot2::ggplot(data = ., aes(x = suit_score, y = predicted)) +
    geom_rug(data = clim_data_africa_presences, aes(x= suit_score, y = pres), sides = "t") + 
    geom_rug(data = clim_data_africa_absences, aes(x= suit_score, y = pres), sides = "b") +
    geom_line() +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.1, fill = "darkred") +
    labs(
      title = "Africa",
      subtitle = paste0("Sensitivity = ", round(sensitivity.africa, 2)),
      x = "MaxEnt suitability score",
      y = "Probability of being recorded"
    ); prob_curve.africa

```

### How well did the model perform in the native range?

```{r eval=FALSE}

reduced_preds = terra::rast("data/reduced_preds.tif")
presPoints = readRDS("data/presPoints.rds")
backPoints = readRDS("data/backPoints.rds")

presPoints.native = presPoints %>%
  dplyr::filter(decimalLongitude > 60)
nrow(presPoints.native )

backPoints.native = backPoints %>%
  dplyr::filter(decimalLongitude > 60)
nrow(backPoints.native )

all.native = rbind(presPoints.native, backPoints.native) %>%
  dplyr::rename(lon = decimalLongitude, lat = decimalLatitude)

NR_extent = sf::st_as_sf(all.native, coords = c("lon", "lat"), crs = 4326)
NR_extent = sf::st_bbox(NR_extent)

native_range_layers = raster::crop(reduced_preds, NR_extent)
#plot(native_range_layers$wc2.1_2.5m_bio_1)

# use the model to predict climate suitability for the native range
predict_maxent_NR = terra::predict(maxentModel1, native_range_layers)

terra::plot(predict_maxent_NR)

terra::writeRaster(predict_maxent_NR, "raster_projections/maxent_projection_native_range.tif",
                   overwrite = TRUE)

##############################################################
# Extract MaxEnt suitability scores at each PRESENCE GPS point
##############################################################
  
# convert presence and background points into vector objects
prespoints.native.vect = terra::vect(
  presPoints.native,
  geom = c("decimalLongitude", "decimalLatitude"),
  crs = "EPSG:4326"
)

backpoints.native.vect = terra::vect(
  backPoints.native,
  geom = c("decimalLongitude", "decimalLatitude"),
  crs = "EPSG:4326"
)

maxent_pres_scores.NR = terra::extract(
      x = predict_maxent_NR, # prediction raster of the native range
      y = prespoints.vect,
      xy = TRUE,
      na.rm = TRUE
    ) %>%
    # Clean df
    dplyr::select(
      lat = y,
      lon = x,
      suit_score = 2 # assign the name of the second column ("median") to "suit_score"
    ) %>%
    tidyr::drop_na(suit_score) %>%
    dplyr::mutate(pres = 1)
  
##############################################################
# Extract MaxEnt suitability scores at each BACKGROUND GPS point
##############################################################
  
maxent_background_scores.NR = terra::extract(
      x = predict_maxent_NR,
      y = backpoints.native.vect,
      xy = TRUE,
      na.rm = TRUE
    ) %>%
    # Clean df
    dplyr::select(
      lat = y,
      lon = x,
      suit_score = 2
    ) %>%
    tidyr::drop_na(suit_score) %>%
    dplyr::mutate(pres = 0)

##############################################################
# Combine dataframes 
##############################################################

clim_data.NR = dplyr::bind_rows(maxent_pres_scores.NR, maxent_background_scores.NR)

mod.NR = glm(pres ~ suit_score, data = clim_data.NR, family = binomial(link = "logit"))
  
print(summary(mod.NR))

# get the beta coefficient
beta_original = coef(mod.NR)["suit_score"]
beta_per_1percent = beta_original * 0.01
odds_ratio_1percent = exp(beta_per_1percent)
cat("Odds ratio for a 1% increase:", odds_ratio_1percent)

# The value of 1.049937 means that the odds of recording the presence of ACP increases by a factor of 1.049937 with a 1% increase in MaxEnt suitability score.

# check model fit
DHARMa::simulateResiduals(fittedModel = mod.NR, plot = TRUE)

car::Anova(mod.NR, test = "LR", type = "II")
# the LR test shows a significant p-value
  
##############################################################
# Plot statistical model  
##############################################################
  
# Extract model predictions 
preds.NR = ggeffects::ggpredict(mod.NR, terms = c("suit_score [0:1 by = 0.01]")) %>%
    as.data.frame() %>%
    dplyr::rename(suit_score = x)
  
clim_data_NR_presences = dplyr::filter(clim_data.NR, pres == 1)
clim_data_NR_absences = dplyr::filter(clim_data.NR, pres == 0)
  
#################################
# sensitivity
#################################

# find the optimal threshold, based on the model
e.NR = dismo::evaluate(p = presPoints.native, a = backPoints.native, model = maxentModel1)

# View various thresholds
thresholds.NR = dismo::threshold(e.NR)
print(thresholds.NR)

# Define a threshold
threshold.NR = 0.29
predicted_presence.NR = ifelse(clim_data.NR$suit_score >= threshold.NR, 1, 0)

# Create confusion matrix
conf_matrix.NR = table(Predicted = predicted_presence.NR, Observed = clim_data.NR$pres)
print(conf_matrix.NR)

# Extract TP and FN
TP.NR = conf_matrix.NR["1", "1"]
FN.NR = conf_matrix.NR["0", "1"]

# Calculate sensitivity
sensitivity.NR = TP.NR / (TP.NR + FN.NR)
cat("Sensitivity:", sensitivity.NR, "\n")

# Plot model predictions 
prob_curve.NR = preds.NR %>%
    ggplot2::ggplot(data = ., aes(x = suit_score, y = predicted)) +
    geom_rug(data = clim_data_NR_presences, aes(x= suit_score, y = pres), sides = "t") + 
    geom_rug(data = clim_data_NR_absences, aes(x= suit_score, y = pres), sides = "b") +
    geom_line() +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.1, fill = "forestgreen") +
    labs(
      title = "Native Range",
      subtitle = paste0("Sensitivity = ", round(sensitivity.NR, 2)),
      x = "MaxEnt suitability score",
      y = "Probability of being recorded"
    ); prob_curve.NR

```

## Exploring other projection areas

Let's plot prediction maps for Tanzania and South Africa specifically, as there are currently surveys being undertaken to survey and monitor ACP in these countries. These maps could help in targeting areas for field work, thereby saving time and effort.

Tanzania:

```{r eval=FALSE}

# read in our predictors again
reduced_preds = terra::rast("data/reduced_preds.tif")

tanzania_ext = rnaturalearth::ne_countries(scale = "medium",
                                           returnclass = "sf") %>%
  dplyr::filter(name %in% c("Tanzania"))

tanzania_ext = sf::st_set_crs(tanzania_ext, 4326)

# Mask reduced set of WORLDCLIM layers to the extent of TZ
tanz_map = raster::crop(reduced_preds, tanzania_ext)

# Extract MaxEnt predictions for Tanzania
predict_maxent_tanz = terra::predict(maxentModel1, tanz_map)
terra::plot(predict_maxent_tanz)

predict_maxent_tanz = terra::mask(predict_maxent_tanz, tanzania_ext)

presPoints = readRDS("data/presPoints.rds")

# Plot publication-quality figure  

tanzania = plot.predmap(raster = predict_maxent_tanz, EXT = tanzania_ext, title = "Tanzania")

```

South Africa:

```{r eval=FALSE}

southafrica_ext = rnaturalearth::ne_countries(scale = "medium",
                                           returnclass = "sf") %>%
  dplyr::filter(name %in% c("South Africa"))

southafrica_ext = sf::st_set_crs(southafrica_ext, 4326)

# Mask reduced set of WORLDCLIM layers to the extent of South Africa
sa_map = raster::crop(reduced_preds, southafrica_ext)

# Extract MaxEnt predictions for Tanzania
predict_maxent_sa = terra::predict(maxentModel1, sa_map)
predict_maxent_sa = terra::mask(predict_maxent_sa, southafrica_ext)
terra::plot(predict_maxent_sa)

provincial_borders = ne_states(country = "South Africa", returnclass = "sf")

# Plot publication-quality figure  

southAfrica = plot.predmap(raster = predict_maxent_sa, EXT = southafrica_ext, 
                           title = "South Africa", arrow.pos = "br", scale.pos = "br") 

southAfrica = southAfrica + 
  # add provincial borders
  geom_sf(data = provincial_borders, fill = NA, color = "black") +
  # modify the lat lon display
  coord_sf(
  xlim = c(16, 33),      
  ylim = c(-35, -22),    
  crs = 4326,           
  expand = FALSE
) 

southAfrica 

```

We can even have a look at Reunion Island, where ACP has been reported recently.

```{r eval=FALSE}
# potentially run this if the function can't find Réunion because of the é
# Sys.setlocale("LC_ALL", "en_US.UTF-8")

reunion_ext = rnaturalearth::ne_states(country = "France", returnclass = "sf") %>%
  dplyr::filter(grepl("Réunion", name, ignore.case = TRUE))

reunion_ext = sf::st_set_crs(reunion_ext, 4326)

# Mask reduced set of WORLDCLIM layers to the extent of Reunion
reunion_map = raster::crop(reduced_preds, reunion_ext)

# Extract MaxEnt predictions for Reunion
predict_maxent_reunion = terra::predict(maxentModel1, reunion_map)
predict_maxent_reunion = terra::mask(predict_maxent_reunion, reunion_ext)
terra::plot(predict_maxent_reunion)

presPoints = readRDS("data/presPoints.rds")

# Plot publication-quality figure   

reunion = plot.predmap(raster = predict_maxent_reunion, EXT =  reunion_ext, title = "Reunion")
reunion

```

## Model assessment

Something interesting to explore is manually excluding occurrence points from a desired area, creating a MaxEnt model without them, and then having a look at how accurate the model is in predicting those areas as being suitable. This approach offers a form of independent validation, which one doesn't often see in the literature.

As an exploratory exercise, let's exclude all the GPS points from South America, and re-run the MaxEnt model. We'll then validate the model using these independent points.

```{r eval=FALSE}

# these were our original presence and background points
presPoints = readRDS("data/presPoints.rds")
backPoints = readRDS("data/backPoints.rds")

# read in our predictors again
reduced_preds = terra::rast("data/reduced_preds.tif")

# let's filter our presence and background points such that we remove all records from South America
# woSA = without South America
presPoints.woSA = presPoints %>%
  dplyr::filter(!(decimalLatitude >= -56 & decimalLatitude <= 12 &
           decimalLongitude >= -81 & decimalLongitude <= -34))

backPoints.woSA = backPoints %>%
  dplyr::filter(!(decimalLatitude >= -56 & decimalLatitude <= 12 &
           decimalLongitude >= -81 & decimalLongitude <= -34))

# compare the number of rows, and see that we have removed 143 presence points
nrow(presPoints)
nrow(presPoints.woSA)

# here we've removed 1222 background points
nrow(backPoints)
nrow(backPoints.woSA)

# plot on a world map to confirm
world_map = rnaturalearth::ne_countries(
  scale = "medium", 
  returnclass = "sf"
) 

# Plot GPS points on world map to check our locality data is correct 
ggplot() +
  # Add raster layer of world map 
  geom_sf(data = world_map, alpha = 0.5) +
  # Add GPS points 
  geom_point(
    data = backPoints.woSA, 
    size = 0.2, colour = "darkred",
    aes(x = decimalLongitude, y = decimalLatitude)
  )  +
  geom_point(
    data = presPoints.woSA, 
    size = 1, colour = "forestgreen",
    aes(x = decimalLongitude, y = decimalLatitude)
  )  +
  # Set world map CRS 
  coord_sf(
    crs = 4326,
    ylim = c(-58, 90),  # Clip below -52 degrees latitude
    expand = FALSE
  ) + 
  xlab("Longitude") + 
  ylab("Latitude")

# bind the presence and absence data together into one data frame
maxent_data.woSA = dplyr::bind_rows(presPoints.woSA, backPoints.woSA) %>% 
  dplyr::select(-c(species, decimalLatitude, decimalLongitude)) 
rownames(maxent_data.woSA) = NULL
head(maxent_data.woSA)
nrow(maxent_data.woSA)

# Create a vector containing 0 (indicating background points) and 1 (indicating presence points)
maxent_vector.woSA = c(
  replicate(nrow(presPoints.woSA), "1"),
  replicate(nrow(backPoints.woSA), "0")
) 

# FIT MAXENT MODEL
maxentModel.woSA = dismo::maxent(
  x = maxent_data.woSA,
  p = maxent_vector.woSA,
  path = here::here("models/maxent_H3.woSA/"),
  replicates = 10,
  args = c(
    # Insert the optimal RM value here
    'betamultiplier=3.0',
    # Turn these on/off to change FC combinations 
    # - To only use quadratic features, turn all to false except quadratic
    'linear=false',
    'quadratic=false',
    'product=false',
    'threshold=false',
    'hinge=true',
    # Don't change anything from here down 
    'threads=2',
    'doclamp=true',
    'fadebyclamping=true',
    'responsecurves=true',
    'jackknife=true',
    'askoverwrite=false',
    'responsecurves=true',
    'writemess=true',
    'writeplotdata=true',
    'writebackgroundpredictions=true'
  )
)

# compare the response plots of the model with and without South American presence points
dismo::response(maxentModel.woSA, expand = TRUE)
dismo::response(maxentModel1, expand = TRUE)

```


Now let's project this new model onto Africa, and compare it to the model where we included South American points.

```{r eval=FALSE}

africa_ext = rnaturalearth::ne_countries(scale = "medium", returnclass = "sf") %>%
  dplyr::filter(continent == "Africa")

# filter to get points just in Africa
species_afr = presPoints %>%
  dplyr::filter(decimalLongitude > -26, decimalLatitude < 55, decimalLatitude > -48, decimalLatitude < 40)

africa_ext = sf::st_set_crs(africa_ext, 4326)

africa_env_layers = terra::rast("africa_env_layers/africa_env_layers.tif")

# use the model to predict climate suitability for Africa
predict_maxent_africa.woSA = terra::predict(maxentModel.woSA, africa_env_layers)
predict_maxent_africa.woSA = terra::mask(predict_maxent_africa.woSA, africa_ext)

terra::plot(predict_maxent_africa.woSA)

terra::writeRaster(predict_maxent_africa.woSA, "raster_projections/maxent_projection_africa.woSA.tif",
                   overwrite = TRUE)

# in case you need to read this back into memory
predict_maxent_africa.woSA = terra::rast("raster_projections/maxent_projection_africa.woSA.tif")

# Plot the map

# remember the function your created earlier? You can use it again here:

africa.woSA = plot.predmap(raster = predict_maxent_africa.woSA, EXT = africa_ext)

# let's add our GPS points to it, and crop it slightly
africa.woSA = africa.woSA +
  geom_point(
    data = species_afr,
    size = 1,
    aes(x = decimalLongitude, y = decimalLatitude)
  )  +
  # Crops map to just the geographic extent of Africa
  coord_sf(
    xlim = c(-25, 55),
    ylim = c(-35, 38),
    crs = 4326,
    expand = FALSE
  )

# save the figure
# ggsave("figs/africa_map.woSA.png", plot = africa.woSA, width = 6, height = 6, dpi = 350)

# save in a format that can be opened in Google Earth as a KML layer
africa_kml.woSA = terra::project(predict_maxent_africa.woSA, "EPSG:4326")
africa_kml.woSA = raster::raster(africa_kml.woSA)
raster::KML(africa_kml.woSA, "figures/africa.woSA", maxpixels=5000000, overwrite = TRUE)

# plot them alongside each other using the gridExtra package
# africa_projections = gridExtra::grid.arrange(africa, africa.woSA, nrow = 1) %>%
  # ggsave("figs/africa_maps.png", plot = ., width = 8, height = 8, dpi = 350)

```

Note the differences in suitable area prediction when the South American points were excluded from the model - particularly along the eastern coastline of South Africa.

Let's restrict our Africa maps to show only the areas that have climatic suitability predictions of 0.75 and above, and calculate the difference in suitable area between the two. We will then validate the second MaxEnt model using the independent South American points.

```{r eval=T}

predict_maxent_africa = terra::rast("raster_projections/maxent_projection_africa.tif")
predict_maxent_africa.woSA = terra::rast("raster_projections/maxent_projection_africa.woSA.tif")

predict_maxent_africa_thresh0.75 = predict_maxent_africa > 0.75
terra::plot(predict_maxent_africa_thresh0.75)

predict_maxent_africa.woSA_thresh0.75 = predict_maxent_africa.woSA > 0.75
terra::plot(predict_maxent_africa.woSA_thresh0.75)

# get the difference between the two rasters
diff_raster = predict_maxent_africa_thresh0.75 - predict_maxent_africa.woSA_thresh0.75
terra::plot(diff_raster, col = c("darkred", "grey85", "darkgreen"))

# calculate km area difference
area_raster = terra::cellSize(diff_raster, unit = "km")
total_area_diff_km2 = terra::global(area_raster * (diff_raster != 0), sum, na.rm = TRUE)$sum
total_area_diff_km2 

# if you want to rather use ggplot, first convert the raster to a dataframe
diff_df = as.data.frame(diff_raster, xy = TRUE, na.rm = TRUE)
head(diff_df)

diffafrica.map = ggplot(diff_df) +
  geom_tile(aes(x = x, y = y, fill = factor(maxent))) + 
  scale_fill_manual(
    values = c("1" = "darkgreen", "0" = "grey95", "-1" = "darkred"),
    name = "Difference",
    labels = c("-1" = "Decrease", "0" = "No Change", "1" = "Increase")
  ) +
  coord_sf(expand = FALSE) +
  labs(
    title = "Difference in Suitability: Full model - model w/o S America",
    x = "Longitude",
    y = "Latitude"
  ) +
  # Crops map to just the geographic extent of Africa
  coord_sf(
    xlim = c(-25, 55),
    ylim = c(-35, 38),
    crs = 4326,
    expand = FALSE
  ) +
  # Create title for the legend
  theme(legend.position = "right") +
  # Add scale bar to bottom-right of map
 ggspatial::annotation_scale(
    location = "bl",          # 'bl' = bottom left
    style = "ticks",
    width_hint = 0.2
  ) +
  # Add north arrow
  ggspatial::annotation_north_arrow(
    location = "bl",
    which_north = "true",
    pad_x = unit(0.175, "in"),
    pad_y = unit(0.3, "in"),
    style = ggspatial::north_arrow_fancy_orienteering
  ) +
  # Change appearance of the legend
  guides(
    fill = guide_colorbar(ticks = FALSE)
  ) ;diffafrica.map

# ggsave("figures/africa_diff.png", plot = diffafrica.map, width = 6, height = 6, dpi = 350)

```

In the above plot, the green areas show where the full model (including South American points) predicted suitable areas (> 0.75) that the model without those points did not, while the red areas show the converse. Grey areas denote no difference between the models.

Let's use the South American points now to validate our latest MaxEnt model. First, we need to create objects that contain presence and background points for South America, and a raster layer with our predictor variables for the continent.

```{r eval=FALSE}

# Since we already had these presence and background points in our original datasets, let's create subsets that contain just those South American points

presPoints.SAm = presPoints %>%
  dplyr::filter(decimalLatitude >= -56 & decimalLatitude <= 12 &
           decimalLongitude >= -81 & decimalLongitude <= -34)

backPoints.SAm = backPoints %>%
  dplyr::filter(decimalLatitude >= -56 & decimalLatitude <= 12 &
           decimalLongitude >= -81 & decimalLongitude <= -34)

# plot on a world map to confirm
world_map = rnaturalearth::ne_countries(
  scale = "medium", 
  returnclass = "sf"
) 

# Plot GPS points on world map to check our locality data is correct 
ggplot() +
  # Add raster layer of world map 
  geom_sf(data = world_map, alpha = 0.5) +
  # Add GPS points 
  geom_point(
    data = backPoints.SAm, 
    size = 0.2, colour = "darkred",
    aes(x = decimalLongitude, y = decimalLatitude)
  )  +
  geom_point(
    data = presPoints.SAm, 
    size = 1, colour = "forestgreen",
    aes(x = decimalLongitude, y = decimalLatitude)
  )  +
  # Set world map CRS 
  coord_sf(
    crs = 4326,
    ylim = c(-58, 90),  # Clip below -52 degrees latitude
    expand = FALSE
  ) + 
  xlab("Longitude") + 
  ylab("Latitude")

# Now let's crop our predictor raster to South America

# read in our predictors again
reduced_preds = terra::rast("data/reduced_preds.tif")

southAmerica_ext = rnaturalearth::ne_countries(scale = "medium",
                                           returnclass = "sf") %>%
  dplyr::filter(continent %in% c("South America"))

southAmerica_ext = sf::st_set_crs(southAmerica_ext, 4326)

# Mask reduced set of WORLDCLIM layers to the extent of South America
southAmerica_map = raster::crop(reduced_preds, southAmerica_ext)

# Get MaxEnt prediction for South America
predict_maxent_SAm = terra::predict(maxentModel.woSA, southAmerica_map)
predict_maxent_SAm = terra::mask(predict_maxent_SAm, southAmerica_ext)
terra::plot(predict_maxent_SAm)

```

Now we need to extract the MaxEnt suitability score at each presence and background point in South America

```{r eval=FALSE}

prespoints.vect = terra::vect(
  presPoints.SAm,
  geom = c("decimalLongitude", "decimalLatitude"),
  crs = "EPSG:4326"
)

backpoints.vect = terra::vect(
  backPoints.SAm,
  geom = c("decimalLongitude", "decimalLatitude"),
  crs = "EPSG:4326"
)
  
##############################################################
# Extract MaxEnt suitability scores at each PRESENCE GPS point
##############################################################
  
maxent_pres_scores = terra::extract(
      x = predict_maxent_SAm, # prediction raster of South America
      y = prespoints.vect,
      xy = TRUE,
      na.rm = TRUE
    ) %>%
    # Clean df
    dplyr::select(
      lat = y,
      lon = x,
      suit_score = 2 # assign the name of the second column ("median") to "suit_score"
    ) %>%
    tidyr::drop_na(suit_score) %>%
    dplyr::mutate(pres = 1)
  
##############################################################
# Extract MaxEnt suitability scores at each BACKGROUND GPS point
##############################################################
  
maxent_background_scores = terra::extract(
      x = predict_maxent_SAm,
      y = backpoints.vect,
      xy = TRUE,
      na.rm = TRUE
    ) %>%
    # Clean df
    dplyr::select(
      lat = y,
      lon = x,
      suit_score = 2
    ) %>%
    tidyr::drop_na(suit_score) %>%
    dplyr::mutate(pres = 0)
  
# Combine dataframes 
clim_data.SAm = dplyr::bind_rows(maxent_pres_scores, maxent_background_scores)

```

Now that we have predicted MaxEnt scores for each presence and background point in South America, we can run a GLM to see whether the presence of ACP can be reliably predicted given a suitability score generated by the model we built (which excluded South American points). In other words, can the model we created distinguish between presence and background points for ACP?

```{r eval=FALSE}

#############################################################
# Fit statistical model  
##############################################################
  
mod.SAm = glm(pres ~ suit_score, data = clim_data.SAm, family = binomial(link = "logit"))
  
print(summary(mod.SAm))

# get the beta coefficient
beta_original = coef(mod.SAm)["suit_score"]
beta_per_1percent = beta_original * 0.01
odds_ratio_1percent = exp(beta_per_1percent)
cat("Odds ratio for a 1% increase:", odds_ratio_1percent)

# The value of 1.005653 means that the odds of recording the presence of ACP increases by a factor of 1.005653 with a 1% increase in MaxEnt suitability score.

# check model fit
DHARMa::simulateResiduals(fittedModel = mod.SAm, plot = TRUE)

car::Anova(mod.SAm, test = "LR", type = "II")
# the LR p-value here is not significant
  
##############################################################
# Plot statistical model  
##############################################################
  
# Extract model predictions 
preds = ggeffects::ggpredict(mod.SAm, terms = c("suit_score [0:1 by = 0.01]")) %>%
    as.data.frame() %>%
    dplyr::rename(suit_score = x)
  
clim_data_presences = dplyr::filter(clim_data.SAm, pres == 1)
clim_data_absences = dplyr::filter(clim_data.SAm, pres == 0)
  
###############################
# Sensitivity score
###############################

# find the optimal threshold, based on the model
e.SAm = evaluate(p = presPoints.SAm, a = backPoints.SAm, model = maxentModel.woSA)

# View various thresholds
thresholds.SAm = dismo::threshold(e.SAm)
print(thresholds.SAm)

# Define a threshold
threshold.SAm = 0.43
predicted_presence.SAm = ifelse(clim_data.SAm$suit_score >= threshold.SAm, 1, 0)

# Create confusion matrix
conf_matrix.SAm = table(Predicted = predicted_presence.SAm, Observed = clim_data.SAm$pres)
print(conf_matrix.SAm)

# Extract TP and FN
TP.SAm = conf_matrix.SAm["1", "1"]
FN.SAm = conf_matrix.SAm["0", "1"]

# Calculate sensitivity
sensitivity.SAm = TP.SAm / (TP.SAm + FN.SAm)
cat("Sensitivity:", sensitivity.SAm, "\n")

# Plot model predictions 
prob_curve_SAm = preds %>%
    ggplot2::ggplot(data = ., aes(x = suit_score, y = predicted)) +
    geom_rug(data = clim_data_presences, aes(x= suit_score, y = pres), sides = "t") + 
    geom_rug(data = clim_data_absences, aes(x= suit_score, y = pres), sides = "b") +
    geom_line() +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.1, fill = "steelblue") +
    labs(
      title = "South America",
      subtitle = paste0("Sensitivity = ", round(sensitivity.SAm, 2)),
      x = "MaxEnt suitability score",
      y = "Probability of being recorded"
    ); prob_curve_SAm

```

```{r, eval=FALSE}

gridExtra::grid.arrange(prob_curve.africa, prob_curve_SAm, prob_curve.NR, nrow = 1)

```